Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-08 paper/main.tex
unifying bounds in the figures and the big table of the appendix

% Comment for anonymization
% Also look for "SHORT" comments
% \def\fullversion{}

@@

%SHORT
\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
\captionsetup[figure]{format=myformat}
\captionsetup[table]{format=myformat}
\renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

%SHORT
\lstset{basicstyle=\tt\scriptsize}
\abovedisplayskip=3pt
\belowdisplayskip=3pt
\abovedisplayshortskip=3pt
\belowdisplayshortskip=3pt
%SHORT

@@

\titlebanner{Draft -- Do not distribute}        % These are ignored unless
\preprintfooter{Draft}   % 'preprint' option specified.

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 3 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.

@@

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, we use a cost metric that counts cost $1$ for assignments
and cost $0$ for all other operations.  Sometimes we also use the
\emph{ticks metric} if we want to discuss features like returning
resources.  Of course, the examples can also be analyzed with any other
cost metric.

We assume that free variables in the code snippets are the inputs to
the program. Some of the examples contain constants on which the
computed bound depends.  These constants are randomly chosen to
present an example but the analysis works for other constants as well.
Note however that it is sometimes crucial that constants are positive
(or negative) and that the same constant at different places.

@@

$3|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t13}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100 && *){
    y=y-100;
  }
}
   \end{lstlisting}

$12|[n,0]| + 0.01|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
   \caption{Amortization and Compositionality (a)}
  \label{fig:cat1a}
\end{figure*}

@@

$4|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t07}
    \end{center}
  \end{minipage}%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x>y) {
    x=x-1;
    x=x+1000;
    y=y+1000;
  }
  while (y>0) {
    y=y-1;
  }
  while (x<0) {
    x=x+1;
  }
   \end{lstlisting}

$1004|[y,x]|+|[x,0]|+|[0,y]|$
\\[.7\baselineskip]
      {\bf t28}
    \end{center}
  \end{minipage}%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (y>=0);
  while (x-y>0) {
    x=x-1;
    x=x-y;
    z=y;
    while (z>0) {
      z=z-1;
    }
  }
   \end{lstlisting}

$3|[0,x]|$
\\[.7\baselineskip]
      {\bf t15}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (y>=0);
  while (x-y>0) {
    x=x-1;
    x=x-y;
    z=y;
    z=z+y;
    z=z+100;
    while (z>0) {
      z=z-1;
    }
  }
   \end{lstlisting}

$1+2|[0,x]|+103|[y,x]|$
\\[.7\baselineskip]
      {\bf t16}
    \end{center}
  \end{minipage}

   \caption{Amortization and Compositionality (b)}
  \label{fig:cat1b}
\end{figure*}

@@

$52 + |[-1,i]| + |[0,k]|$
\\[.7\baselineskip]
      {\bf t19}
    \end{center}
  \end{minipage}%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x<y) {
    x=x+1;
  }
  while (y<x) {
    y=y+1;
  }
   \end{lstlisting}

@@

$4|[0,x]|+4|[0,y]|$
\\[.7\baselineskip]
      {\bf t30}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    if (n>0 && *) {
      n=n-1;
      flag=1;
    } else
      flag=0;
  }
   \end{lstlisting}

$2 + 2|[0, n]|$
\\[.7\baselineskip]
      {\bf t47}
    \end{center}
  \end{minipage}

   \caption{Amortization and Compositionality (c)}
  \label{fig:cat1c}
\end{figure*}

@@

Figures~\ref{fig:cat1a}, \ref{fig:cat1b}, and~\ref{fig:cat1c} show
code snippets that need amortization are compositionality to obtain a
whole program bound.

@@

Example \emph{t28} is particularly interesting.  In the first loop we
decrement the size $|[y,x]|$.  However, we also shift the interval
$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
that this does not change the size of the interval and computes the
tight loop bound $3|[y,x]|$.  The additional two loops are in the
program to show that the size tracking in the first loop works
accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
the worst case (if $x$ and $y$ are negative).

Sometimes we need some assumptions on the inputs in order to derive a
bound.  Example \emph{t15} is such a case.  We assume here that the
input variable $y$ is non-negative and write \code{assert(y>=0)}.  The
semantic of \code{assert} is that it has no effect if the assertion is
true and that the program is terminated without further cost
otherwise.  If we enter the loop then we know that $x>0$ and we can
obtain constant potential from the assignment \code{x=x-1}.  After the
assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
can share the potential $3|[0,x]|$ before the assignment \code{x=x-y}
between $3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we
derive a tight linear bound.

Example \emph{t16} is an extension of example \emph{t15}. We again
assume that $y$ is non-negative and use the same mechanism to iterate
the outer loop as in \emph{t15}.  In the inner loop, we also count the
variable $z$ down to zero and perform $|[0,z]|$ interations.  However,
instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
computes the tight linear bound $1+2|[0,x]|+103|[y,x]|$.  The
assignment of potential to the size interval $|[y,x]|$ in instead of
$|[0,x]|$ is a random choice of the LP solver.  Other optimal
solutions of the linear program that the analyzer gernerates for
\emph{t15} would result in bounds such as $1+105|[0,x]|$ or
$1+6|[0,x]|+99|[y,x]|$.  In fact, the $105$ potential units can be
shared arbitrarily between the interval sizes $|[y,x]|$ and $|[0,x]|$
as long as more than $1$ unit is assigned to $|[0,x]|$.  Since we know
that, $y\geq 0$ the choice of the LP solver is optimal but this seems
to be a coincidence.

Example \emph{t19} demonstrates another nice property that demonstates
the compositionaliy of the analysis.  The program consists of two
loops that decrement a variable $i$.  In the first loop, $i$ is
decremented down to 100 and in the second loop $i$ is increment
further down to $-1$.  However, between the loops we assign
$i=i+k+50$.  So in total the program performs $52 + |[-1,i]| +
|[0,k]|$ increments.  Our analysis finds this tight bound because our
amortized analysis naturally takes into account the relation between
the two loops.  Other techniques might derive a more conservative
bound such as $52 + |[-1,i]| + |[0,k]| + |[100,i]|$.

@@

At first sight, example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Our analysis
infers the tight bound $4|[0,x]|+4|[0,y]|$.

Example \emph{t47} demonstrates how we can use integers as Booleans to
amortize the cost of loops that depend on boolean flags.  The outer
loop is executed as long as the variable flag is ``true'', that is,
\code{flag>0}.  Inside the loop, there is a conditional that either
(if $n>0$) decrements $n$ and assigns \code{flag=1}, or (if $n\geq0$)
leaves n unchanged and assigns \code{flag=0}.  The analyzer computes
the tight bound $2 + 2|[0, n]|$.  The potential in the loop invariant
is $|[0,\mathit{flag}]| + 2|[0, n]|$.  In the \emph{then} branch of
the conditional, we use the potential $2|[0, n]|$ and the fact that
$n>0$.  In the \emph{else} branch, we use the potential
$|[0,\mathit{flag}]|$ and the fact that $\mathit{flag}=1$.

@@

$1 + |[0, n]|$
\\[.7\baselineskip]
      {\bf nested\_single}
    \end{center}
  \end{minipage}

   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09}} (a)
  \label{fig:cat2a}
\end{figure*}

@@

$1 + |[0,n]|$
\\[.7\baselineskip]
      {\bf sequential\_single}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  x=0; y=0;
  while (x<n) {
    if (y<m)
      y=y+1;
    else
      x=x+1;
  }
   \end{lstlisting}
$2 + |[0, m]| + |[0, n]|$
\\[.7\baselineskip]
      {\bf simple\_multiple}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  x=0;
  while (x<n) {
    if (*)
      x=x+1;
    else 
      x=x+1;
  }
   \end{lstlisting}

$1 + |[0,n]|$
\\[.7\baselineskip]
      {\bf simple\_single}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  x=0; y=0;
  while (*) {
    if (x<N) {
      x=x+1; y=y+1;
    } else if (y<M ) {
      x=x+1; y=y+1;
    } else
      break;
  }
   \end{lstlisting}

$2 + 2 |[0, M]| + 2 |[0, N]|$
\\[.7\baselineskip]
      {\bf simple\_single\_2}
    \end{center}
  \end{minipage}

   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b)}
  \label{fig:cat2b}
\end{figure*}

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert n>0;
  assert m>0;
  va = n; vb = 0;
  while (va>0 && *) {
    if (vb<m) { 
      vb=vb+1; 
      va=va-1;
    } else {
      vb=vb-1;
      vb=0;
    }
  }
   \end{lstlisting}
$2 + 4|[0, n]|$
\\[.7\baselineskip]
      {\bf fig4\_2}
    \end{center}
  \end{minipage}
%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0<m);
  i = n;
  while (i>0 && *) {
    if (i<m)
      i=i-1;
    else
      i=i-m;
  }
   \end{lstlisting}
$|[0, n]|$
\\[.7\baselineskip]
      {\bf fig4\_4}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0 < m < n);
  i=m;
  while (0<i<n) {
    if (dir==fwd) i++;
    else i--;
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf fig4\_5}
    \end{center}
  \end{minipage}

   \caption{Examples from [GulwaniPLDI09]}
  \label{fig:cat2c}
\end{figure*}

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  i=0;
  while (i<n) {
    j=i+1;
    while (j<n) {
      if (*) {
        tick(1);
        j=j-1; n=n-1;
      }
      j=j+1;
    }
    i=i+1;
  }
   \end{lstlisting}
$|[0, n]| \text{ ticks}$
\\[.7\baselineskip]
      {\bf ex1}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0 && m>0) {
    n--; m--;
    while (nondet()) {
      n--; m++;
    }
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf ex2}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0) {
    t = x;
    n=n-1;
    while (n>0) {
      if (*) break;
      n=n-1;
    }
  }
   \end{lstlisting}
$2 |[0, n]|$
\\[.7\baselineskip]
      {\bf ex3}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    flag=0;
    while (n>0 && *) {
      n=n-1;       
      flag=1;
    }
  }
   \end{lstlisting}
$2 + 3|[0, n]|$
\\[.7\baselineskip]
      {\bf ex4}
    \end{center}
  \end{minipage}

   \caption{Examples from [GulwaniPLDI10]}
  \label{fig:cat3a}
\end{figure*}

@@

Our analyzer can derive almost all linear bounds for programs that
have been described as challenges in the literature on bound
generation.  We found only three programs with a linear bound for
which our analyzer could not find a tight bound.  The first one
(\emph{fig4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
  reasoning} to derive a bound.  The other two, \emph{ex2} and
\emph{ex5} (see [GulwaniPLDI10]), from [GulwaniPLDI10] were described
to have a linear bound but seem to be non-terminating.

@@

Example \emph{simple\_single\_2} uses conditionals and a \code{break}
statement to control loop iterations.  If $x<N$ then variables $x$ and
$y$ are incremented.  Otherwise, if $y<M$ then the same increment is
executed.  If $y\geq M$ and $x\geq N$ then the loop is terminated with
a break.  Our tool computes the bound $2 + 2 |[0, M]| + 2 |[0, N]|$.
This bound is tight in the sense that there are inputs (such as $M =
-100$ and $N = 100$) for which the bound precisely describes the
execution cost.  However, SPEED can compute the more precise bound
$\max(N,M)$.  We currently cannot express this bound in our system.

Example \emph{fig4\_2} from [GulwaniPLDI09] is quite involved.
Amortized reasoning helps to understand how we derive the bound $4 +
4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
constant potential to pay for the two first assignments.  We use the
remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
loop invariant.  In the \emph{if branch} of the conditional, we use
the constant potential $4$ of the invariant to pay for the two
assignments ($2$ units) and the potential of $|[0,vb]|$ ($2$ units).
Since we also know that $|[0, n]|>0$ we obtain constant potential $4$
and establish the loop invariant again.  In the \emph{else} branch, we
use the potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$
potential units to pay for the assignment.  Since we have $4$ constant
potential units left after the loop, we can subtract them from the
initial potential $6 + 4|[0, n]|$.

In example \emph{fig4\_4} it is essential that $m$ is positive.
That ensures that we can obtain constant potential for the interval
size $|[0,i]|$ in the \emph{else} branch of the conditional since
$|[0,i]|$ decreases.  Example \emph{fig4\_5} is one of the three
examples that we can not handle automatically.  The execution is
bounded because the boolean value of the test \code{dir==fwd} does not
change during the iteration of the loop.  As a result, the variable
$i$ is either counted down to $0$ or up to $n$.  Our tool cannot
handle example \emph{fig4\_5} because we don't do path sensitive
reasoning.  Note however that it would be more efficient to move the
test \code{dir==fwd} outside of the loop (this would be also done by
an optimizing compiler).  The resulting program could be analyzer by our
tool.

>>>>

% Comment for anonymization
% Also look for "SHORT" comments
\def\fullversion{}

@@

%SHORT
% \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
% \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
% \captionsetup[figure]{format=myformat}
% \captionsetup[table]{format=myformat}
% \renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

%SHORT
% \lstset{basicstyle=\tt\scriptsize}
% \abovedisplayskip=3pt
% \belowdisplayskip=3pt
% \abovedisplayshortskip=3pt
% \belowdisplayshortskip=3pt
%SHORT

@@

\titlebanner{\iffull{Extended Version}{}}%{Draft -- Do not distribute}        % These are ignored unless
\preprintfooter{\iffull{Extended Version\hspace{-.8cm}}{}}%{Draft}   % 'preprint' option specified.

@@

We have evaluated the automatic analysis with system code and examples
from the literature.  \iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with the tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicability and
expressiveness of the quantitative Hoare logic.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds the maximal element in an array, and a linear
bound that amortizes the cost of $n$ successive increments to a binary
counter.

@@

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, and to compare our analysis with existing tools, we use a
cost metric that counts the number of loop iterations and function
calls.  This is the only cost metric supported by tools like
KoAT~\cite{BrockschmidtEFFG14} and LOOPUS~\cite{SinnZV14}.  Sometimes
we also use the \emph{tick metric} if we want to discuss features such as
resource restitution.  The tick metric assigns cost $n$ to the
statement \code{tick(n)} and cost $0$ to all other statements. Of
course, the examples can also be analyzed with any other cost metric
in our system.

We assume that free variables in the code snippets are the inputs of
the program. Some of the examples contain constants on which the
computed bound depends.  These constants are randomly chosen to
present an example but the analysis works for other constants as well.
Note however that it is sometimes crucial that constants are positive
(or negative) or that other relations hold.

\pref{tab:eval} contains the details of our comparative evaluation.

@@

$2|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t13}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100 && *){
    y=y-100;
  }
}
   \end{lstlisting}

$11|[n,0]| + 0.01|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
   \caption{Amortization and Compositionality (a).}
  \label{fig:cat1a}
\end{figure*}

@@

$1 + 3|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t07}
    \end{center}
  \end{minipage}%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x>y) {
    x=x-1;
    x=x+1000;
    y=y+1000;
  }
  while (y>0) {
    y=y-1;
  }
  while (x<0) {
    x=x+1;
  }
   \end{lstlisting}

$1002|[y,x]|+|[x,0]|+|[0,y]|$
\\[.7\baselineskip]
      {\bf t28}
    \end{center}
  \end{minipage}%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (y>=0);
  while (x-y>0) {
    x=x-1;
    x=x-y;
    z=y;
    while (z>0) {
      z=z-1;
    }
  }
   \end{lstlisting}

$|[0,x]|$
\\[.7\baselineskip]
      {\bf t15}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (y>=0);
  while (x-y>0) {
    x=x-1;
    x=x-y;
    z=y;
    z=z+y;
    z=z+100;
    while (z>0) {
      z=z-1;
    }
  }
   \end{lstlisting}

$101|[y,x]|$
\\[.7\baselineskip]
      {\bf t16}
    \end{center}
  \end{minipage}

   \caption{Amortization and Compositionality (b).}
  \label{fig:cat1b}
\end{figure*}

@@

$50 + |[-1,i]| + |[0,k]|$
\\[.7\baselineskip]
      {\bf t19}
    \end{center}
  \end{minipage}%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x<y) {
    x=x+1;
  }
  while (y<x) {
    y=y+1;
  }
   \end{lstlisting}

@@

$|[0,x]|+|[0,y]|$
\\[.7\baselineskip]
      {\bf t30}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    if (n>0 && *) {
      n=n-1;
      flag=1;
    } else
      flag=0;
  }
   \end{lstlisting}

$1 + |[0, n]|$
\\[.7\baselineskip]
      {\bf t47}
    \end{center}
  \end{minipage}

   \caption{Amortization and Compositionality (c).}
  \label{fig:cat1c}
\end{figure*}

@@

Figures~\ref{fig:cat1a}, \ref{fig:cat1b}, and~\ref{fig:cat1c} show
code snippets that need amortization and compositionality to obtain a
whole program bound.

@@

Example \emph{t28} is particularly interesting.  In the first loop we
decrement the size $|[y,x]|$.  However, we also shift the interval
$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
that this does not change the size of the interval and computes the
tight loop bound $|[y,x]|$.  The additional two loops are in the
program to show that the size tracking in the first loop works
accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
the worst case (if $x$ and $y$ are negative).

Sometimes we need some assumptions on the inputs in order to derive a
bound.  Example \emph{t15} is such a case.  We assume here that the
input variable $y$ is non-negative and write \code{assert(y>=0)}.  The
semantic of \code{assert} is that it has no effect if the assertion is
true and that the program is terminated without further cost
otherwise.  If we enter the loop then we know that $x>0$ and we can
obtain constant potential from the assignment \code{x=x-1}.  After the
assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
can share the potential $|[0,x]|$ before the assignment \code{x=x-y}
between $|[0,x]|$ and $|[0,y]|$ after the assignment.  In this way, we
derive a tight linear bound.

Example \emph{t16} is an extension of example \emph{t15}. We again
assume that $y$ is non-negative and use the same mechanism to iterate
the outer loop as in \emph{t15}.  In the inner loop, we also count the
variable $z$ down to zero and perform $|[0,z]|$ iterations.  However,
instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
computes the linear bound $101|[0,x]|$.  The
assignment of potential to the size interval $|[0,x]|$ in instead of
$|[y,x]|$ is a random choice of the LP solver.

Example \emph{t19} consists of two loops that decrement a variable
$i$.  In the first loop, $i$ is decremented down to 100 and in the
second loop $i$ is increment further down to $-1$.  However, between
the loops we assign $i=i+k+50$.  So in total the program performs $50
+ |[-1,i]| + |[0,k]|$ iterations.  Our analysis finds this tight bound
because our amortized analysis naturally takes into account the
relation between the two loops.  Techniques that do not use
amortization derive a more conservative bound such as $50 + |[-1,i]| +
|[0,k]| + |[100,i]|$.

@@

At first sight, example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Our analysis
infers the tight bound $|[0,x]|+|[0,y]|$.

Example \emph{t47} demonstrates how we can use integers as Booleans to
amortize the cost of loops that depend on boolean flags.  The outer
loop is executed as long as the variable flag is ``true'', that is,
\code{flag>0}.  Inside the loop, there is a conditional that either
(if $n>0$) decrements $n$ and assigns \code{flag=1}, or (if $n\geq0$)
leaves n unchanged and assigns \code{flag=0}.  The analyzer computes
the tight bound $1 + |[0, n]|$.  The potential in the loop invariant
is $|[0,\mathit{flag}]| + |[0, n]|$.  In the \emph{then} branch of
the conditional, we use the potential $|[0, n]|$ and the fact that
$n>0$.  In the \emph{else} branch, we use the potential
$|[0,\mathit{flag}]|$ and the fact that $\mathit{flag}=1$.

@@

$|[0, n]|$
\\[.7\baselineskip]
      {\bf nested\_single}
    \end{center}
  \end{minipage}

   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (a).}
  \label{fig:cat2a}
\end{figure*}

@@

$|[0,n]|$
\\[.7\baselineskip]
      {\bf sequential\_single}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  x=0; y=0;
  while (x<n) {
    if (y<m)
      y=y+1;
    else
      x=x+1;
  }
   \end{lstlisting}
$|[0, m]| + |[0, n]|$
\\[.7\baselineskip]
      {\bf simple\_multiple}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  x=0;
  while (x<n) {
    if (*)
      x=x+1;
    else 
      x=x+1;
  }
   \end{lstlisting}

$|[0,n]|$
\\[.7\baselineskip]
      {\bf simple\_single}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  x=0; y=0;
  while (*) {
    if (x<N) {
      x=x+1; y=y+1;
    } else if (y<M ) {
      x=x+1; y=y+1;
    } else
      break;
  }
   \end{lstlisting}

$|[0, M]| + |[0, N]|$
\\[.7\baselineskip]
      {\bf simple\_single\_2}
    \end{center}
  \end{minipage}

   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b).}
  \label{fig:cat2b}
\end{figure*}

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert n>0;
  assert m>0;
  va = n; vb = 0;
  while (va>0 && *) {
    if (vb<m) { 
      vb=vb+1; 
      va=va-1;
    } else {
      vb=vb-1;
      vb=0;
    }
  }
   \end{lstlisting}
$1 + 2|[0, n]|$
\\[.7\baselineskip]
      {\bf fig4\_2}
    \end{center}
  \end{minipage}
%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0<m);
  i = n;
  while (i>0 && *) {
    if (i<m)
      i=i-1;
    else
      i=i-m;
  }
   \end{lstlisting}
$|[0, n]|$
\\[.7\baselineskip]
      {\bf fig4\_4}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0 < m < n);
  i=m;
  while (0<i<n) {
    if (dir==fwd) i++;
    else i--;
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf fig4\_5}
    \end{center}
  \end{minipage}

   \caption{Examples from~\cite{GulwaniJK09}.}
  \label{fig:cat2c}
\end{figure*}

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  i=0;
  while (i<n) {
    j=i+1;
    while (j<n) {
      if (*) {
        tick(1);
        j=j-1; n=n-1;
      }
      j=j+1;
    }
    i=i+1;
  }
   \end{lstlisting}
$|[0, n]| \text{ ticks}$
\\[.7\baselineskip]
      {\bf ex1}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0 && m>0) {
    n--; m--;
    while (nondet()) {
      n--; m++;
    };
    tick(1);
  }
   \end{lstlisting}
$|[0,n]|$ ticks
\\[.7\baselineskip]
      {\bf ex2}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0) {
    t = x;
    n=n-1;
    while (n>0) {
      if (*) break;
      n=n-1;
    }
  }
   \end{lstlisting}
$|[0, n]|$
\\[.7\baselineskip]
      {\bf ex3}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    flag=0;
    while (n>0 && *) {
      n=n-1;       
      flag=1;
    }
  }
   \end{lstlisting}
$1 + 2|[0, n]|$
\\[.7\baselineskip]
      {\bf ex4}
    \end{center}
  \end{minipage}

   \caption{Examples from~\cite{GulwaniZ10}.}
  \label{fig:cat3a}
\end{figure*}

@@

Our analyzer can derive almost all linear bounds for programs that
have been described as challenges in the literature on bound
generation.  We found only one program with a linear bound for
which our analyzer could not find a tight bound:  Example
(\emph{fig4\_5}) from~~\cite{GulwaniJK09} requires \emph{path-sensitive
  reasoning} to derive a bound.

@@

Example \emph{simple\_single\_2} uses conditionals and a \code{break}
statement to control loop iterations.  If $x<N$ then variables $x$ and
$y$ are incremented.  Otherwise, if $y<M$ then the same increment is
executed.  If $y\geq M$ and $x\geq N$ then the loop is terminated with
a break.  Our tool computes the bound $|[0, M]| + |[0, N]|$.
This bound is tight in the sense that there are inputs (such as $M =
-100$ and $N = 100$) for which the bound precisely describes the
execution cost.  However, SPEED can compute the more precise bound
$\max(N,M)$.  We currently cannot express this bound in our system.

Example \emph{fig4\_2} from~\cite{GulwaniJK09} is quite involved.
Amortized reasoning helps to understand how we derive the bound $1 +
2|[0, n]|$.  We start with potential $1 + 2|[0, n]|$ and use the fact
that $vb=0$ to establish the potential $1 + 2|[0, n]| + |[0,vb]|$ that
serves as a loop invariant.  In the \emph{if branch} of the
conditional, we use the constant potential $1$ of the invariant to pay
for the potential of $|[0,vb]|$.  Since we also know that $|[0, n]|>0$
we obtain constant potential $2$ that we use to pay for the loop
iteration ($1$ unit) and to establish the loop invariant again ($1$
unit).  In the \emph{else} branch, we use the potential $|[0,vb]|$
and the fact $vb>0$ to obtain $1$ potential units to pay for the
loop iteration.

In example \emph{fig4\_4} it is essential that $m$ is positive.  That
ensures that we can obtain constant potential for the interval size
$|[0,i]|$ in the \emph{else} branch of the conditional since $|[0,i]|$
decreases.  Example \emph{fig4\_5} is an examples that we can not
handle automatically.  The execution is bounded because the boolean
value of the test \code{dir==fwd} does not change during the iteration
of the loop.  As a result, the variable $i$ is either counted down to
$0$ or up to $n$.  Our tool cannot handle example \emph{fig4\_5}
because we don't do path sensitive reasoning.  Note however that it
would be more efficient to move the test \code{dir==fwd} outside of
the loop (this would be also done by an optimizing compiler).  The
resulting program can then be analyzed by our tool.

Example \emph{ex1} from~\cite{GulwaniZ10} specifically focuses on the
code in the \emph{if} statement.  So we use the \emph{tick metric} and
insert \code{tick(1)} inside the if statement to derive a bound on
the number of times the code in the if statement is executed.  Note
that we cannot derive a bound for the whole program since the outer
loop is executed a quadratic number of times.  Nevertheless it is
straightforward to derive a bound on the number of ticks using the
amortized approach: In the if statement we know that $n>0$ and assign
\code{n=n-1}.  So we can use the potential of the interval size $|[0,n]|$
to pay for the tick.
